---
title: "Data Cleaning Process"
author: "Chenming Ran"
date: "4/26/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r package, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(tidymodels)
library(haven)
```

In general, a gravity model assumes that the volume of trade between any two economies will be directly proportional to the product of their economic masses that are commonly measured by GDP or GNP. For predicting aggregate bilateral trade, the aggregate GDP value is the suitable proxy in term of cross-country time-series data. While Circlaeys et al.(2017) used  nominal GDP, this study uses trade quantity and real GDP for estimation, which is universally recognized as a more accurate measure of production output 

Although “Tradhist” from Centre d'Etudes Prospectives et d'Informations Internationales (CEPII) is the original database used by Circlaeys et al. (2017), the data is only available up to the year 2014. Different from Circlaeys et al. (2017) analyzing bilateral trade flow from 2009 to 2014, I analyze bilateral trade from 1995 to 2018, so key indicators of Bilateral trade flow and GDP will be recreated and updated. Time-invariant indicators, including weighted distances and other control variables, will be adopted from "Tradehist".

# Bilateral Trade Values
Other than "Tradehist", BACI" is the major bilateral trade database provided by CEPII. This database provides yearly data on bilateral trade flows at the product level. Products are identified using the Harmonized System (HS), a standard nomenclature for international trade, Ths HS was revised in 1992, 1996, 2002, 2007, 2012 and 2017. For this study, I use HS92 version that has data vailable from 1995 to 2018, the bilateral trade value is obtained by country and year.

Due to the extremely large data volume, I first append all HS92 country data from 1995 to 2007.
```{r append Bilateral trade files 1, message=FALSE, warning=FALSE}
library(foreign)
#link to the seperate files of yearly data
filenames <- list.files(path="BAC/BAC1",pattern="*.csv")
#link to the folder containing data files
fullpath=file.path("BAC/BAC1/",filenames)
#append
BAC_1 <- Reduce(rbind, lapply(fullpath, read.csv))
```

Then, I collapsed bilateral trade value by year and country pair (one origin country pairs with one destination country).
```{r Bilateral Trade Updated 1, message=FALSE, warning=FALSE}
BAC_complete_1 <- BAC_1 %>%
  filter(t,i,j,v) %>%
  group_by(t,i,j) %>%
  summarize(value = sum(v)) %>%
  mutate(FLOW = value * 1000) %>%   #change thousand U.S. dollars to U.S Dollars
  select(-value) %>%
  rename("year" = "t", "country_code_o" = "i", "country_code_d" = "j") 

summary(duplicated (BAC_complete_1))  #The data is correct and has no duplicates

#save a subfile 
write_csv(BAC_complete_1, "BAC_complete_1.csv") 

rm(BAC_1)
```

The processes of appending and collapsing are iterated for HS92 country data from 2008 to 2018.
```{r append Bilateral trade files 2, message=FALSE, warning=FALSE}
library(foreign)
file_2 <- list.files(path="BAC/BAC2",pattern="*.csv")
path_2 = file.path("BAC/BAC2/",file_2)
BAC_2 <- Reduce(rbind, lapply(path_2, read.csv))
```

```{r Bilateral Trade Updated 2, message=FALSE, warning=FALSE}
BAC_complete_2 <- BAC_2 %>%
  filter(t,i,j,v) %>%
  group_by(t,i,j) %>%
  summarize(value = sum(v)) %>%
  mutate(FLOW = value * 1000) %>%   #change thousand U.S. dollars to U.S Dollars
  select(-value) %>%
  rename("year" = "t", "country_code_o" = "i", "country_code_d" = "j") 

summary(duplicated (BAC_complete_2))  #The data is correct and has no duplicates

#save a subset
write_csv(BAC_complete_2, "BAC_complete_2.csv")

rm(BAC_2)
```

Then I combine the two dataframe into a complete bilateral trade dataframe, and then use inflation rate (Deflator) to obtained the  real bilateral trade value. 
```{r all BAC files, message=FALSE, warning=FALSE}
BAC_complete = Reduce(rbind, lapply(c("BAC_complete_1.csv", "BAC_complete_2.csv"), read.csv))
#save a file of Bilateral trade value

write_dta(BAC_complete, "BAC_complete.dta")

rm(BAC_complete_1,BAC_complete_2)
```

The the complete bilateral trade dataframe are merged with country codes for trade orgin and destination countries, which prepares later merges of GDP, population and other varibales.
```{r country code, message=FALSE, warning=FALSE}
Country <- read_csv("BAC/country_codes_V201901.csv") %>%
  select(CountryCode, iso3) %>%
  rename("country_code_o" = "CountryCode", "iso_o" = "iso3") %>%
  mutate(iso_d = iso_o, country_code_d = country_code_o)

Country_1 <- Country %>%
  select (country_code_o, iso_o)

Country_2 <- Country %>%
  select (country_code_d, iso_d)

```

```{r merge BAC with country_code, message=FALSE, warning=FALSE}
Trade_1 <- left_join(x = BAC_complete,
                         y = Country_1,
                         by = "country_code_o") 
#"Trade" is bilateral trade dateframe
Trade <- left_join (x = Trade_1, y = Country_2, by = "country_code_d" )
rm(Trade_1, Country, Country_1, Country_2, BAC_complete)
```

# Real GDP
The project gathers GDP (current U.S.$) and inflation rate from the World Development Indicators database, then the real GDP is calculated as Nominal GDP / (1 +inflation Rate). The following is the process of obtaining real GDP. The the real GDP is matched with its targeted year and country. I create GDP_1 as a subset to match with trade originated countries, and GDP_2 to match with trade destination countries. 

```{r deflate real GDP, message=FALSE, warning=FALSE}
library("readxl")
GDP_s <- read_excel("GDP $US.xlsx") %>%
  pivot_longer(c(-CountryCode,-CountryName), names_to = "year", values_to = "GDP") %>%
  arrange(CountryCode, year) 

GDP_s$GDP = as.numeric(GDP_s$GDP)

Deflator <- read_excel("Deflator.xlsx") %>%
  pivot_longer(c(-CountryCode,-CountryName), names_to = "year", values_to = "Inflation") %>%
  arrange(CountryCode, year) 

Deflator$Inflation = as.numeric(Deflator$Inflation)
Deflator[is.na(Deflator)] <- 0  #if not inflation rate, keep the nominal value

GDP_update <- left_join(x= GDP_s, y= Deflator, by =c("CountryCode", "CountryName", "year")) %>%
  rename("iso_o" = "CountryCode", "country" = "CountryName") %>%
  mutate(iso_d = iso_o, deflate = 1 + Inflation/100) %>%
  mutate(Real_GDP = GDP/deflate) %>%
  mutate(GDP_o = Real_GDP, GDP_d = Real_GDP)

## Dataframe GDP is the complete dataset for real GDP per capita
GDP <- GDP_update[, c("country","year","iso_o", "iso_d", "GDP_o", "GDP_d", "GDP", "Inflation")] 

## subset of GDP for late data process
GDP_1 <- GDP %>%
  select(country, year,iso_o, GDP_o) %>%
  rename("country_o" = "country")

GDP_2 <- GDP %>%
  select(country, year, iso_d, GDP_d) %>%
  rename("country_d" = "country")

rm(GDP_update, GDP_s)
```

# Population

Population is a control variable commonly used in the framework of the Gravity Model and was used by Circlaeys et al (2017). Therefore, this study also incorportations country total population data from United Nations Population Division  — World Population Prospect from 1995 to 2018. I create pop_1 as a subset to match with trade origin countries, and pop_2 to match with trade destination countries. 
 pivot_longer(c(-CountryCode,-CountryName), names_to = "year", values_to = "pop") %>%
   arrange(CountryCode, year) %>%
  rename("iso_o" = "CountryCode", "country" = "CountryName") %>%
  mutate(pop_o = pop, pop_d = pop)

```{r population, message=FALSE, warning=FALSE}
pop <- read_excel("population.xlsx")  %>%
  mutate_all(as.character) %>%
  pivot_longer(c(-CountryCode,-CountryName), names_to = "year", values_to = "pop") %>%
  arrange(CountryCode, year) %>%
  rename("iso_o" = "CountryCode", "country" = "CountryName") %>%
  mutate(pop_o = pop, pop_d = pop, iso_d = iso_o)

pop$pop = as.numeric(pop$pop)

## subset of population for late data process
pop_1 <- pop %>%
  select(country, year,iso_o, pop_o) %>%
  rename("country_o" = "country")

pop_2 <- pop %>%
  select(country, year, iso_d, pop_d) %>%
  rename("country_d" = "country")

rm(pop)
```

Then together I creates a complete panel dataset of bilateral trade value, real GDP and population "New_Trade_Data" by merging those dataframes all together.
```{r merge GDP with Trade, message=FALSE, warning=FALSE}
Trade$year = as.character(Trade$year)
GDP_up1<- left_join(x = Trade,y = GDP_1, by = c("year","iso_o")) 

GDP_Trade_Complete<- left_join(x = GDP_up1, y = GDP_2, by = c("year","iso_d")) 

pop_up1 <- left_join (x= GDP_Trade_Complete, y = pop_1, by = c("year", "iso_o", "country_o"))

GDP_Pop_Trade_Complete <- left_join (x = pop_up1, y = pop_2, by = c("year", "iso_d", "country_d") )

#sort columns and sort rows by year and country
New_Trade_Data <- GDP_Pop_Trade_Complete [c(1, 2, 3, 5, 6, 7, 9, 4, 8, 10, 11, 12)]  %>%
  arrange(year, country_code_o, country_code_d)

New_Trade_Data$pop_o = as.numeric(New_Trade_Data$pop_o)
New_Trade_Data$pop_d = as.numeric(New_Trade_Data$pop_d)

#save the new data
write_dta(New_Trade_Data, "New_Trade_Data.dta")
rm(GDP_up1, GDP_1, GDP_2, GDP, Trade, GDP_Trade_Complete, GDP_Pop_Trade_Complete, Deflator, pop_1, pop_2, pop_up1)
```

# Distance and Other controll Variables

The gravity model also incorporates distance between trade origin and destination. Circlaeys et al.(2017) uses 'great circle distance’ between two countries from the trade_hist data base. This indicator has been updated to 'weighted great circle distance' by the author of original "Tradehist" database, and it is a bilateral time-invariant indicator. Thus this study include this distance proxy of each country pair.

The bilateral time-invariant indicators from "Tradehist" database also include other controls like language, sharing a common border that are used by Circlaey et al.(2017). Togther with weighted great circle distance, those indicators are equivalent as those from the database "GeoDsit", another CEPII database incorporates country-specific geographical variables. I incorporate data for distance and bilateral time-invariant controls from "GeoDist" 

OECD member and GATT member were also used by Circlaeys et al.(2017), thus included from country time-invariant dataset. To note, the status of OECD and GATT is not completely time-invariant because countries joined OECD and WTO in different years. The two organizations also have new membership from 2014 to 2018, which I updated by hand in the excel sheet beforehand.

However, different from Circlaeys et al.(2017), this study does not consider value of exports from the origin country as a predictor because of possible colinearity.

```{r time-invariant indicators, message=FALSE, warning=FALSE}
binvar <- read_stata ("GeoDist.dta") %>%
  select(iso_o, iso_d, distw, comlang_ethno, contig)  %>%
  rename("comlang" = "comlang_ethno")
cinvar <- read_excel("country_timeinvariant.xlsx") 

ci_1 <- cinvar %>%
  rename ("OECD_o" = "OECD", "iso_o" = "iso", "GATT_o" = "GATT") 
  
ci_2 <- cinvar %>%
  rename ("OECD_d" = "OECD", "iso_d" = "iso", "GATT_d" = "GATT") 

invar_1 <- left_join(x = ci_1, y = binvar, by = c("iso_o")) 
invar <- left_join(x = invar_1, y = ci_2, by = c("iso_d", "year")) 

#Replace missing of OECD and GATT with 0 because it's determined if a country is a member or not
invar$OECD_d[is.na(invar$OECD_d)] <- 0  
invar$GATT_d[is.na(invar$GATT_d)] <- 0  

#complete df "tinvar" of time-invariant variable
tinvar <- invar [c(2, 1, 5, 6, 7, 8, 3, 4, 9, 10)] 

tinvar$year = as.character(tinvar$year)

rm(binvar, cinvar, invar, invar_1, ci_1, ci_2)
```

Finally I merge time-invariant indicators with the previous dataframe that contains country time-series bilateral trade value, GDP, and population. As a result, I obtain a brandnew bilateral trade dataset "bitrade" for gravity model analysis

```{r merge new trade data with time_invariant data, message=FALSE, warning=FALSE}
new_bitrade <- left_join(x = New_Trade_Data, y = tinvar, by = c("year","iso_o", "iso_d")) 

#save the dataset
write_dta(new_bitrade, "new_bitrade.dta")

rm(tinvar, New_Trade_Data)
```

## Reference

[1] G.Gaulier and S. Zignago, "International Trade Database at the Product-Level." CEPII Working Paper, 2010.

[2] M. Fouquin and J. Hugot, “Two centuries of bilateral trade and gravity data: 1827-2014.” CEPII Working Paper, 2016.

[3] S. Circlaeys, C. Kanitkar, and D. Kumazawa, “Bilateral Trade Flow Prediction”, available at: http://cs229.stanford.edu/proj2017/finalreports/5240224.pdf, 2017

[4] T. Mayer and S. Zignago, "Notes on CEPII’s distances measures : the GeoDist Database." CEPII Working Paper, 2011.

[5] "United Nations Population Division. World Population Prospects: Total Population" World Bank Database, available at: https://data.worldbank.org/indicator/sp.pop.totl, 2019

[6] "World Development Indicators: GDP (current US$)" The World Bank Group, available at: https://data.worldbank.org/indicator/NY.GDP.MKTP.CD, 2019

[7] "World Bank National Account Data: Inflation, GDP deflator (annual %)" The World Bank Group, available at: https://data.worldbank.org/indicator/NY.GDP.DEFL.KD.ZG, 2019